---
title: "Identify lung-specific transcriptomic features"
output: html_document
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F)
library(ArchR)
library(dplyr)
library(tidyr)
library(colorRamp2)
library(scales)
library(htmltools)
library(ComplexHeatmap)
library(cowplot)
set.seed(1)
ct_order <-
  c(
  "Naive_B",
  "Memory_B",
  "NK",
  "CD8_T",
  "CD4_T",
  "Th17",
  "Treg",
  "Other"
)
name_order <- c("Memory_B", "Naive_B", "NK", "CD4_T", "CD8_T")
pal_atac <- readRDS("~/projects/Lung_scMultiomics_paper/data/color_codes.RDS")
pal_rna <- pal_atac[2:8]
source("code/make_plots.R")

```

# Differential gene expression analyses with DESeq2

log(Counts) ~ factor(submission_date) + Race + Age + Sex + Tissue

## Summarizing DE genes by selecting p-value cutoffs

A table of cell counts by tissue and cell-type. 
```{r}
seRNA <- readRDS("~/cluster/projects/u19_multiomics/analyses/u19_seRNA_summarized_experiment.RDS") 
colData(seRNA)$CellType <- 
factor(seRNA$majority_voting,
       labels = c("NK",
                  "NK", "Other", 
                  "Other", "Other",
                  "Memory_B","Naive_B", "Other",
                  "Treg", "CD4_T", 
                  "CD8_T", "Th17")
       )
colData(seRNA)$CellType <- factor(
    colData(seRNA)$CellType,
    levels = 
            c("Other", "Treg", "Th17", "CD4_T",
              "CD8_T", "NK", "Memory_B", "Naive_B"))
table(seRNA$CellType, seRNA$tissue.ident)
```

A barplot for number of DE genes detected for each cell type except for Th17 and Treg, due to low number of cells in spleen.

```{r, fig.width = 10}
tests <- readRDS("output/u19_multiomics/output/u19_rna/bulk_cross_tissue_DESeq2_full.RDS")
tests <- lapply(names(tests), function(i) {
  data.frame(tests[[i]], cluster = gsub("-", "_", i)) %>%
    filter(!is.na(log2FoldChange))
})

DGE_tbl <- do.call(rbind, tests)

gginput <-
    DGE_tbl %>% 
      mutate(avg_log2FC = log2FoldChange) %>%
      filter(cluster != "Other" & padj <= 0.05) %>%
      mutate(group = case_when(avg_log2FC >= 1~ "Lung Up-regulated", 
                                 avg_log2FC <= -1 ~ "Spleen Up-regulated")) 

totalCounts <- gginput %>% group_by(cluster, group) %>% dplyr::summarise(n = n())

ggplot(
  totalCounts %>% 
    filter(!cluster %in% c("Th17", "Treg") &
             grepl("Up-regulated", group)),
  aes(x = factor(cluster, levels = ct_order[1:5]),
      y = n,
      fill = group)) + 
  theme_bw(base_size  = 15) + 
  geom_col(width = 0.4) +
  theme_ArchR(baseSize = 15, legendTextSize = 15) + 
  #geom_label(size = 2, position = position_dodge2()) + 
  xlab("") + ylab("#DE genes with at least two fold change") + 
  ggtitle("Majority of DE genes were detected from memory B cells for both tissues")
```

### A combined volcano plot for DE genes across cell types
```{r, warning = F}
library(ggrepel)
selected <- ct_order[1:5]
ggplot(
  gginput %>% 
    filter(cluster %in% selected),
  aes(
    x = avg_log2FC ,
    y = -log10(pvalue),
    color = cluster,
  )
) +
  geom_point(size = 0.5) +
  scale_color_manual(values = c(pal_rna)) + 
  geom_point(
    data =
      gginput %>%
      filter(group == "Not Significant") %>%
      filter(cluster %in% selected),
    aes(x = avg_log2FC, y = -log10(pvalue)),
    color = alpha("grey", 0.7),
    size = 0.5
  ) +
  geom_text_repel(
    data = gginput %>% filter(cluster %in% selected),
    aes(label = ifelse(abs(avg_log2FC) >= 0.58, gene, "")),
    color = "grey30",
    max.overlaps = 20,
    nudge_x = 0.1, nudge_y = 0.1,
    size =2.5) +
  facet_wrap(~ factor(cluster, levels = ct_order[1:5]), nrow = 2) +
  theme_pub() + xlim(c(-5, 10)) + 
  theme(legend.position = "none")

```

### DE statistics for the *GNLY* example

```{r}
gginput %>% filter(gene == "GNLY")
```


```{r, eval = F}
# comparing RNA counts between tissue for GNLY
rna_counts <- readRDS("~/cluster/projects/u19_multiomics/output/u19_rna/pseudobulk_RNA_counts.RDS")
samples = unlist(lapply(colnames(rna_counts), function(i) {
  paste(strsplit(i, "-")[[1]][1:2], collapse = "-")}))
sample_cov <- fread("data/sample_covariates.txt")
rna_metadata <- sample_cov[match(samples, sample_cov$core_ID), 1:8]
rna_metadata$CellType = unlist(lapply(colnames(rna_counts), function(i) {
  paste(strsplit(i, "-")[[1]][-2:-1], collapse = "_")})
  )

rna_counts <- rna_counts[, rna_metadata$CellType %in% ct_order[1:5]]
libSizes <- apply(rna_counts, 2, sum)
rna_norm <- rna_counts/libSizes*(10^6)

rna <- rna_counts[rownames(rna_counts) == "GNLY", ]

gginput <-
  lapply(unique(rna_metadata$CellType), function(i) {
    data.frame(
      Lung = rna[which(rna_metadata$cells== "Lung" & rna_metadata$CellType == i)],
      Spleen = rna[which(rna_metadata$cells== "Spleen" & rna_metadata$CellType == i)],
      CellType = i
    )
  })

gginput <- do.call(rbind, gginput)

ggplot(gginput %>% 
         filter(CellType %in% ct_order[1:5]) %>%
         pivot_longer(!CellType, 
                      names_to = "Tissue",
                      values_to = "Counts"), 
       aes(x = CellType,
           y = log2(Counts),
           fill = Tissue)) + 
  xlab("") + ylab("log2(Counts)") +
  geom_violin() + 
  geom_jitter(position = position_dodge(width = 0.9), size = 0.7) + 
  theme_pub()
```

### GO enrichment for cross-tissue genes

```{r}
# preparing inputs for GO enrichment

# bkgGenes <- unique(tests$Other$gene)
# lung_up <- gginput[gginput$group == "Lung Up-regulated", ]
# genesets <- lapply(split(lung_up, lung_up$cluster), function(i) {i$gene})
# spleen_up <- gginput[gginput$group == "Spleen Up-regulated", ]
# genesets <- lapply(split(spleen_up, spleen_up$cluster), function(i) {i$gene})
# visulize GO enrichment outputs
go_list <- readRDS("output/u19_multiomics/output/GO_enrichment/btw_tissue/DESeq2/lung_upreg_GO_enrichment.RDS")
feature = "Biological"
go_tbl <-
  do.call(rbind, go_list) %>%
    filter(k %in% ct_order[1:5])

toshow <- 
    go_tbl %>% 
      filter(FDR <= 0.05 & grepl(feature, database)) %>%
          group_by(k) %>%
          slice_max(n = 8, order_by = enrichmentRatio) %>%
          pull(description)

ggplot(
    go_tbl %>% 
      filter(grepl(feature, database) & description %in% toshow) %>%
      mutate(FDR_adj = ifelse(FDR == 0, 10^(-10), FDR)),
    aes(y = description,
        x = factor(k, levels = ct_order[1:5]),
        size = enrichmentRatio,
        color = -log10(FDR_adj))) + 
    geom_point() + xlab("") + ylab("") + 
     # scale_fill_gradient(
     #   limits = c(0, 4),
     #   oob = squish) +
    theme_pub() + ggtitle("GO BP enrichment for lung up-regulated genes") + 
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```

```{r}
go_list <- readRDS("output/u19_multiomics/output/GO_enrichment/btw_tissue/DESeq2/spleen_upreg_GO_enrichment.RDS")
feature = "Biological"
go_tbl <-
  do.call(rbind, go_list) %>%
    filter(k %in% ct_order[1:5])

toshow <- 
    go_tbl %>% 
      filter(FDR <= 0.05 & grepl(feature, database)) %>%
          group_by(k) %>%
          slice_max(n = 8, order_by = enrichmentRatio) %>%
          pull(description)

ggplot(
    go_tbl %>% 
      filter(grepl(feature, database) & description %in% toshow) %>%
      mutate(FDR_adj = ifelse(FDR == 0, 10^(-10), FDR)),
    aes(y = description,
        x = factor(k, levels = ct_order[1:5]),
        size = enrichmentRatio,
        color = -log10(FDR_adj))) + 
    geom_point() + xlab("") + ylab("") + 
     # scale_fill_gradient(
     #   limits = c(0, 4),
     #   oob = squish) +
    theme_pub() + ggtitle("GO BP enrichment for spleen up-regulated genes") + 
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```












